---
title: "DepsearcheR - querying syntactically annotated data with R"
author: "Juho Härme"
date: "`r Sys.Date()`"
output:
    gitbook:
        split_by: none
    
---


The DepsearcheR package is a simple utility made for the purpose of using R for
corpus analyses involving the utilization of dependency annotations represented
in some version of the CoNLL format (cf. e.g [here](https://stackoverflow.com/questions/27416164/what-is-conll-data-format)).

# Example data
 

Assume that you have parsed a file like the one provided in the `inst/extdata`
folder of this package (the Finnish wikipedia article for sparrow):

```{r}

library(depsearcheR)
library(readr)
mytext <- readr::read_file(
                           system.file("extdata", 
                                       "varpunen_wikipedia.txt",
                                       package="depsearcheR")
                           )

cat(substr(mytext,1,300))


```

The text has been parsed with [the Finnish dependency parser developed at the
university of Turku](http://turkunlp.github.io/Finnish-dep-parser/) and this
output file is also included in `inst/extdata`. Note that 
the format here is the so called universal dependencies format.
This is what the conll formatted file looks like:


    1	Varpunen	varpunen	NOUN	_	Case=Nom|Number=Sing	8	nsubj:cop	_	_
    2	(	(	PUNCT	_	_	4	punct	_	_
    3	Passer	Passer	PROPN	_	Case=Nom|Number=Sing	4	compound:nn	_	_
    4	domesticus	domesticus	NOUN	_	Case=Nom|Number=Sing	1	appos	_	_
    5	)	)	PUNCT	_	_	4	punct	_	_
    6	on	olla	VERB	_	Mood=Ind|Number=Sing|Person=3|Tense=Pres|VerbForm=Fin|Voice=Act	8	cop	_	_
    7	yleinen	yleinen	ADJ	_	Case=Nom|Degree=Pos|Number=Sing	8	amod	_	_
    8	lintulaji	lintu#laji	NOUN	_	Case=Nom|Number=Sing	0	root	_	_
    9	suuressa	suuri	ADJ	_	Case=Ine|Degree=Pos|Number=Sing	10	amod	_	_
    10	osassa	osa	NOUN	_	Case=Ine|Number=Sing	8	nmod	_	_
    11	Eurooppaa	Eurooppa	PROPN	_	Case=Par|Number=Sing	10	nmod	_	_
    12	ja	ja	CONJ	_	_	11	cc	_	_
    13	Aasiaa	Aasia	PROPN	_	Case=Par|Number=Sing	11	conj	_	_
    14	.	.	PUNCT	_	_	8	punct	_	_

    1	Carolus	Carolus	PROPN	_	Case=Nom|Number=Sing	2	name	_	_
    2	Linnaeus	Linnaeus	PROPN	_	Case=Nom|Number=Sing	3	nsubj	_	_
    3	antoi	antaa	VERB	_	Mood=Ind|Number=Sing|Person=3|Tense=Past|VerbForm=Fin|Voice=Act	0	root	_	_
    4	varpuselle	varpunen	NOUN	_	Case=All|Number=Sing	3	nmod	_	_
    5	aluksi	aluksi	ADV	_	_	3	advmod	_	_
    6	nimen	nimi	NOUN	_	Case=Gen|Number=Sing	7	nmod:poss	_	_
    7	Fringilla	Fringilla	NOUN	_	Case=Ade|Number=Plur	3	nmod	_	_
    8	domestica	domestica	X	_	Foreign=Foreign	3	dobj	_	_
    9	.	.	PUNCT	_	_	3	punct	_	_


Now, let's imagine we have a data set consisting of all the sentences of
the Finnish wikipedia article mentioned above. It could be acquired
as follows:

```{r}


library(dplyr)
library(readr)
sentences <- readr::read_file(
                           system.file("extdata", 
                                       "varpunen.conll",
                                       package="depsearcheR")
                           ) %>% 
        strsplit("\n\n")  %>% 
        unlist



```

The data is also included in the package as a sample vector
called `varpunen_sentences`, so to use that in the following examples, just run


```{r}

sentences <- varpunen_sentences

```


Now, since there is some variation among the different conll formats
and not all conll outputs have the same number of columns, you should
set the columns used in each session as a global option 
via the `options` function. For instance, for the results
of the Stanford parser's 2015 version, we should do:

```{r}
options("conll_cols" = c("tokenid","token","lemma","feat","none", "head", "dep"))
```

Since in the following examples we'll be using the Finnish dep parser, we'll
set it like this:

```{r}
options("conll_cols" = c("tokenid","token","lemma","pos","pos2","feat","head","dep","null1","null2"))
```





# Usage on the sentence level

At the heart of the depsearcheR package is an extremely simple
function called `FilterConllRows`. The idea of this function
is to filter a conll formatted sentence according to some
conditions given by the users. The output of the filter
can then be used for more filtering.


As simple and trivial scenario, imagine that we want to retrieve all the
nouns of a sentence. This could be achieved in the following way
(using the example set of sentences from the previous section):

```{r}

FilterConllRows(sentences[1], "pos", "NOUN")

```

The function can also be used with regular expressions:


```{r}
FilterConllRows(sentences[1], "feat", "Case=Ine", use_regex=T)
```

...conditions with multiple values

```{r}
FilterConllRows(sentences[1], "pos", c("NOUN","ADJ"))
```

or with negative conditions, for instance to get everything
that's not a noun or an adjective:

```{r}
FilterConllRows(sentences[1], "pos", c("NOUN","ADJ"), is_negative=T)
```

You can add more complex conditions by piping the 
reslts in `dplyr` style. For instance, to 
get something that is not a noun but is in the inessive
case:

```{r}

FilterConllRows(sentences[1], "pos", c("NOUN","PROPN"),is_negative=T) %>% 
    FilterConllRows("feat", "Case=Ine", use_regex=T) 
```

This is where the function actually gets useful for
getting information about dependencies.


Now, imagine we want to get all the dependents of a 
finite verb in a sentence. For that, we can use
the values of the columns `head` and `tokenid`.

```{r}


#Get all the finite verbs in the sentence
finverbs <- FilterConllRows(sentences[2], "feat", "VerbForm=Fin", T)
#Get their dependents
deps <- FilterConllRows(sentences[2], "head", finverbs$tokenid)
deps
                         
```

There is actually a shortcut function for this (`GetDeps`). It takes
exactly the same arguments as 'FilterConllRows'.
Hence, the previous example could also be written as:


```{r}


#Get the dependents of finite verbs
deps <- GetDeps(sentences[2],"feat","VerbForm=Fin", T)
deps
                         
```

There's also a convienience function for testing
if a sentence contains a certain dependency relation,
say, if there is at least one adjective governed
by a noun. This function is called `ContainsDepRel`
and it can be used as follows:

```{r}

ContainsDepRel(sentences[1],
             depw=c("pos","ADJ"),
             headw=c("pos","NOUN")
             )


```



# Usage on the dataset level

All the previous examples demonstrate the usage of the simple functions in
this package on the sentence level. What the whole package is actually
for, is, however, querying datastructures containing multiple sentences
and filtering only the ones that are relevant for the user.

It should be emphasized at this point that this package is definitely not
meant for effective data mining on a large scale. The 
functions used are simple, perhaps even trivial, and performance
is the cost. But if you know what you're searching for and 
are not dealing with a terribly large dataset (or have a lot of time)
this can be a useful approach.

## Querying a raw text

The first, although propably not the most common, use case considered
here is where you have a conll formatted text (such as the one from the first section
of this vignette) and you want to make queries to find sentences with certain
characteristics. To do that, we want to have the data formatted
as a vector of sentences.

Since the data in conll format separates sentences with an empty row, 
a vector of sentences can be obtained simply by splitting the file
by two consequtive newlines as was done at the beginning of this
vignette. There is also a convenience function for this, which
takes a filename as a parameter and produces the vector:

```{r}

sentences  <- GetSentencesFromFile("../inst/extdata/varpunen.conll")

```

Now, let's say we want to get all the sentences 

1) which include a finite verb
2) in which the finite verb has a subject dependent
3) in which the subject is a  proper name

<!-- -->

The trick here is to develop custom functions that will take care 
of the filtering and then apply those functions to the vector
containing the sentences. Here's an example function that can be used
for searching for only the kind of results described above (has a finite verb,
the verb has a subject dependent, the subject is a proper name)

```{r}

MyFilterFunction <- function(sentence) {
            my_matches <- 
                GetDeps(sentence,"feat","VerbForm=Fin", T) %>% 
                FilterConllRows("dep","nsubj") %>% 
                FilterConllRows("pos","PROPN") 

            return(nrow(my_matches))
} 


```

There is one requirement to keep in mind when creating 
these filters: every filter function should return
a truthy value based on the matches retrieved. The simplest
way to do this is to return the number of rows in the
tibble returned by the sentence filters.
To run the filter, we can use R's built-in `Filter` function
(note: uppercase F), which takes a function as the first
argument and a vector as the second:

```{r}

matched_sentences <- Filter(MyFilterFunction, sentences)

```

In our case, there was one sentence that matched the query.
To view the sentence as a tibble, we can use the function
`ConllAsTibble`:

```{r}

ConllAsTibble(matched_sentences)


```

Or, if we would like a more human-readable representation of the actual 
sentence, there is a function called `ConllAsSentence`:

```{r}

ConllAsSentence(matched_sentences)

```

Keep in mind, however, that these two functions take *single* sentences
as arguments, not vectors of sentences.

This kind of querying works okay with the kind of toy files used here
as examples. However, if we have a larger text, this really not
an effective way of mining it. However, if you don't mind
the query running for a while, you can, of course, give 
this approach a go. In that case, it might be useful to 
add a progress bar to the function, for instance 
with the `progress_estimated` function from `dplyr`.
Here's an example with the previous function
transformed to an unnamed function inside the call
to `Filter`:

```{r}
library(dplyr)

#initialize the progress bar
p <- progress_estimated(length(sentences))

matched_sentences <- Filter(
            function(sentence) {
                my_matches <- 
                    GetDeps(sentence,"feat","VerbForm=Fin", T) %>% 
                    FilterConllRows("dep","nsubj") %>% 
                    FilterConllRows("pos","PROPN") 
                #Move the progress bar:
                p$tick()$print()
                return(nrow(my_matches))
            }, sentences)

#Remove the progress bar
p$stop()

```

The small wikipedia article doesn't, of course, cause any performance problems,
but with longer texts the progress bar is nice to have. As a larger example, 
consider querying the entire (English translation of the) gospel of John[^john] with 
a similar filter:


```{r}

options("conll_cols" = c("tokenid","token","lemma","feat","none", "head", "dep"))
sentences  <- GetSentencesFromFile("../inst/extdata/gospel_of_john.txt.conll")

#initialize the progress bar
p <- progress_estimated(length(sentences))

matched_sentences <- Filter(
            function(sentence) {
                my_matches <- 
                    GetDeps(sentence,"feat","VBZ", T) %>% 
                    FilterConllRows("dep","nsubj") %>% 
                    FilterConllRows("feat","NN") 
                #Move the progress bar:
                p$tick()$print()
                return(nrow(my_matches))
            }, sentences)

#Remove the progress bar
p$stop()

```

TODO: a function called GetColFromConll etc

On my thinkpad x220 that took around 3 seconds -- still not terribly bad.

I actually tested with the whole new testament, and that took.
That might give you some idea of the possibilities of this utterly simple
library in terms of performance and larger texts.

[^john]: Included, also, as a sample file from the public domain World English
version and parsed with the Stanford coreNLP parser.





## Filtering concordances

Although it is certainly possible to use this library for 
querying entire texts outputted in the conll format, the use case
for which the library was created is when you have a data frame
containing a list of concordances with more information than
just the actual conll-annotated sentence.

The basic idea is to have the data in a data frame / tibble where one row
corresponds to one sentence and there is a column containing the conll
representation of the sentence. In addition, you probably have some variables,
at least metadata an the like. One sample use case is, for instance, if you've queried 
for Twitter data and would like to have each tweet parsed and then use
the conll output of the tweet together with the massive amounts of 
metadata variables provided by the Twitter API for each tweet.

Here's a toy example of such a task.

Let's imagine we want to analyze the usage of the word "referee" in 
recent Uefa champions' league games (recent, because the Twitter api
only let's you search for tweets not older than one week).
In order to run the following you need to have requested 
developer access from Twitter, but never mind,
the data is also available as a small data set accompanying this 
package.

```{r}

#library(rtweet)
#ref <- search_tweets("(#UCL OR #ChampionsLeague OR #UEFAChampionsLeague) referee",  n = 5000)

```

I used the method described in the [webcorpcrawler package](https://github.com/utacorpora/webcorpcrawler/blob/master/docs/add_annotations_example.md)
to add syntactic annotations using Stanford Core-NLP. This gave me the following data set,
available as part of this package.
(I've stripped away most of the props offered by the Twitter API since there is A LOT of them)

```{r}

ucl_ref

```





such as the source of the concordance


Another basic use case is if you have a data set consisting of concordances,
acquired through, for instance, a query to a web corpus




# TODO

TODO: first make a reference data frame with all data as tabbed?
TODO: function to list possible columns and values






