---
title: "DepsearcheR - querying syntactically annotated data with R"
author: "Juho Härme"
date: "`r Sys.Date()`"
output:
    gitbook:
        split_by: none
    
---


The DepsearcheR package is a simple utility made for the purpose of using R for
corpus analyses involving the utilization of dependency annotations represented
in some version of the CoNLL format (cf. e.g [here](https://stackoverflow.com/questions/27416164/what-is-conll-data-format)).

# Example data
 

Assume that you have parsed a file like the one provided in the `inst/extdata`
folder of this package (the Finnish wikipedia article for sparrow):

```{r}

library(depsearcheR)
library(readr)
mytext <- readr::read_file(
                           system.file("extdata", 
                                       "varpunen_wikipedia.txt",
                                       package="depsearcheR")
                           )

cat(substr(mytext,1,300))


```

The text has been parsed with [the Finnish dependency parser developed at the
university of Turku](http://turkunlp.github.io/Finnish-dep-parser/) and this
output file is also included in `inst/extdata`. Note that 
the format here is the so called universal dependencies format.
This is what the conll formatted file looks like:


    1	Varpunen	varpunen	NOUN	_	Case=Nom|Number=Sing	8	nsubj:cop	_	_
    2	(	(	PUNCT	_	_	4	punct	_	_
    3	Passer	Passer	PROPN	_	Case=Nom|Number=Sing	4	compound:nn	_	_
    4	domesticus	domesticus	NOUN	_	Case=Nom|Number=Sing	1	appos	_	_
    5	)	)	PUNCT	_	_	4	punct	_	_
    6	on	olla	VERB	_	Mood=Ind|Number=Sing|Person=3|Tense=Pres|VerbForm=Fin|Voice=Act	8	cop	_	_
    7	yleinen	yleinen	ADJ	_	Case=Nom|Degree=Pos|Number=Sing	8	amod	_	_
    8	lintulaji	lintu#laji	NOUN	_	Case=Nom|Number=Sing	0	root	_	_
    9	suuressa	suuri	ADJ	_	Case=Ine|Degree=Pos|Number=Sing	10	amod	_	_
    10	osassa	osa	NOUN	_	Case=Ine|Number=Sing	8	nmod	_	_
    11	Eurooppaa	Eurooppa	PROPN	_	Case=Par|Number=Sing	10	nmod	_	_
    12	ja	ja	CONJ	_	_	11	cc	_	_
    13	Aasiaa	Aasia	PROPN	_	Case=Par|Number=Sing	11	conj	_	_
    14	.	.	PUNCT	_	_	8	punct	_	_

    1	Carolus	Carolus	PROPN	_	Case=Nom|Number=Sing	2	name	_	_
    2	Linnaeus	Linnaeus	PROPN	_	Case=Nom|Number=Sing	3	nsubj	_	_
    3	antoi	antaa	VERB	_	Mood=Ind|Number=Sing|Person=3|Tense=Past|VerbForm=Fin|Voice=Act	0	root	_	_
    4	varpuselle	varpunen	NOUN	_	Case=All|Number=Sing	3	nmod	_	_
    5	aluksi	aluksi	ADV	_	_	3	advmod	_	_
    6	nimen	nimi	NOUN	_	Case=Gen|Number=Sing	7	nmod:poss	_	_
    7	Fringilla	Fringilla	NOUN	_	Case=Ade|Number=Plur	3	nmod	_	_
    8	domestica	domestica	X	_	Foreign=Foreign	3	dobj	_	_
    9	.	.	PUNCT	_	_	3	punct	_	_


Now, let's imagine we have a data set consisting of all the sentences of
the Finnish wikipedia article mentioned above. It could be acquired
as follows:

```{r}


library(dplyr)
library(readr)
sentences <- readr::read_file(
                           system.file("extdata", 
                                       "varpunen.conll",
                                       package="depsearcheR")
                           ) %>% 
        strsplit("\n\n")  %>% 
        unlist



```

The data is also included in the package as a sample vector
called `varpunen_sentences`, which we will use in the
following examples.


Now, since there is some variation among the different conll formats
and not all conll outputs have the same number of columns, you should
set the columns used in each session as a global option 
via the `options` function. For instance, for the results
of the Stanford parser's 2015 version, we should do:

```{r}
options("conll_cols" = c("tokenid","token","lemma","feat","none", "head", "dep"))
```

Since in the following examples we'll be using the Finnish dep parser, we'll
set it like this:

```{r}
options("conll_cols" = c("tokenid","token","lemma","pos","pos2","feat","head","dep","null1","null2"))
```



# Usage on the sentence level

At the heart of the depsearcheR package is an extremely simple
function called `FilterConllRows`. The idea of this function
is to filter a conll formatted sentence according to some
conditions given by the users. The output of the filter
can then be used for more filtering.


As simple and trivial scenario, imagine that we want to retrieve all the
nouns of a sentence. This could be achieved in the following way
(using the example set of sentences from the previous section):

```{r}

FilterConllRows(varpunen_sentences[1], "pos", "NOUN")

```

The function can also be used with regular expressions:


```{r}
FilterConllRows(varpunen_sentences[1], "feat", "Case=Ine", use_regex=T)
```

...conditions with multiple values

```{r}
FilterConllRows(varpunen_sentences[1], "pos", c("NOUN","ADJ"))
```

or with negative conditions, for instance to get everything
that's not a noun or an adjective (or a punctuation mark):

```{r}
FilterConllRows(varpunen_sentences[1], "pos", c("NOUN","ADJ","PROPN","PUNCT"), is_negative=T)
```

You can add more complex conditions by piping the 
reslts in `dplyr` style. Note that since the philosophy
behind this package relies quite heavily on 
the one behind dplyr (or the `tidyverse` framework in general)
it might be a good idea to get a basic idea of what that
is about (e.g. [here](https://dplyr.tidyverse.org/articles/dplyr.html#piping)).
The very basic thing to note: the ` %>% ` operator 
used in the following examples is called the piping operator (from
the `magrittr` package) and it's a way to pass on a functions
return value to another function.


So, to get, for instance,  something that is not a noun but is in the inessive
case, we might do:

```{r}

FilterConllRows(varpunen_sentences[1], "pos", c("NOUN","PROPN"),is_negative=T) %>% 
    FilterConllRows("feat", "Case=Ine", use_regex=T) 
```

And actually, with the piping style going on, we could start with the sentence:

```{r}

varpunen_sentences[1]  %>% 
    FilterConllRows("pos", c("NOUN","PROPN"),is_negative=T) %>% 
    FilterConllRows("feat", "Case=Ine", use_regex=T) 
```

Stylistically, this looks a little better to me.


This kind of "filtering the filtered" is where the function actually gets
useful for getting information about dependencies.


Now, imagine we want to get all the dependents of a 
finite verb in a sentence. For that, we can use
the values of the columns `head` and `tokenid`.

```{r}


#Get all the finite verbs in the sentence
finverbs <- FilterConllRows(varpunen_sentences[2], "feat", "VerbForm=Fin", T)
#Get their dependents
deps <- FilterConllRows(varpunen_sentences[2], "head", finverbs$tokenid)
deps
                         
```

There is actually a shortcut function for getting the dependents of a word.
It's called `GetDeps` and it takes as it's arguments 

- a word (=a row of a tibble) or multiple words (a tibble) that are the heads
  we're looking at
- an unfiltered sentence (as a tibble)

So, the previous example could be written as:

```{r}
varpunen_sentences[2] %>% 
    FilterConllRows("feat", "VerbForm=Fin", T) %>%  # returns one row, passed on to the GetDeps function
    GetDeps(varpunen_sentences[2])  # note: the original sentence is the second argument
```

There is also the `GetHeads` function for doing this the other way
around: if we want to, say, find all the heads of (common) nouns, 
we could do it as follows.


```{r}
varpunen_sentences[2] %>% 
    FilterConllRows("pos", "NOUN") %>%  
    GetHeads(varpunen_sentences[2])  
```


Note that we can use the output of `GetDeps` and `GetHeads` for
further filtering, for instance to get only the kind of 
dependents for finite verbs that are *not* nouns (or punctuation marks):

```{r}
varpunen_sentences[2] %>% 
    FilterConllRows("feat", "VerbForm=Fin", T) %>%  
    GetDeps(varpunen_sentences[2])   %>% 
    FilterConllRows("pos", c("PROPN", "NOUN", "PUNCT"), is_negative=T)
```



# Usage on the dataset level

All the previous examples demonstrate the usage of the simple functions in
this package on the sentence level. What the whole package is actually
for, is, however, querying datastructures containing multiple sentences
and filtering only the ones that are relevant for the user.

It should be emphasized at this point that this package is definitely not
meant for effective data mining on a large scale. The 
functions used are simple, perhaps even trivial, and performance
is the cost. But if you know what you're searching for and 
are not dealing with a terribly large dataset (or have a lot of time)
this can be a useful approach.

## Querying a raw text

The first, although propably not the most common, use case considered
here is where you have a conll formatted text (such as the one from the first section
of this vignette) and you want to make queries to find sentences with certain
characteristics. To do that, we want to have the data formatted
as a vector of sentences.

Since the data in conll format separates sentences with an empty row, 
a vector of sentences can be obtained simply by splitting the file
by two consequtive newlines as was done at the beginning of this
vignette. There is also a convenience function for this, which
takes a filename as a parameter and produces the vector:

```{r}

sentences  <- GetSentencesFromFile("../inst/extdata/varpunen.conll")

```

Now, let's say we want to get all the sentences 

1) which include a finite verb
2) in which the finite verb has a subject dependent
3) in which the subject is a  proper name

<!-- -->

The trick here is to develop custom functions that will take care 
of the filtering and then apply those functions to the vector
containing the sentences. Here's an example function that can be used
for searching for only the kind of results described above (has a finite verb,
the verb has a subject dependent, the subject is a proper name)

```{r}

MyFilterFunction <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat","VerbForm=Fin", T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("pos","PROPN")  %>% 
        return
} 


```

There is one requirement to keep in mind when creating these filters: every
filter must take a single sentence as its only argument and return the filtered
matches so that they can be further processed. Befor the return statement
you can have as complex or as simple a set of code as you wish.
The filters should be used with the `ApplyConllFilter` function of this library
in the following manner:


```{r}

options("conll_cols" = c("tokenid","token","lemma","pos","pos2","feat","head","dep","null1","null2"))
matched_sentences <- ApplyConllFilter(varpunen_sentences, MyFilterFunction)

```

By default `ApplyConllFilter` just returns the sentences of the input
vector that matched the specified filter. 
In our case, there was one sentence that matched the query.
To view the sentence as a tibble, we can use the function
`ConllAsTibble`:

```{r}

ConllAsTibble(matched_sentences)


```

Or, if we would like a more human-readable representation of the actual 
sentence, there is a function called `ConllAsSentence`:

```{r}

ConllAsSentence(matched_sentences)

```

Keep in mind, however, that these two functions take *single* sentences
as arguments, not vectors of sentences.

If we want different kind of output from `ApplyConllFilter`, we can
specify this via the `return_type` parameter. This parameter
is a string defaulting to "raw" (the kind of output we got above, i.e. just
a filtered vector of conll-formatted sentences). Other
possible values are:

- `matches`: returns all the words (rows) that match as a single tibble. 
- `both`: returns both the matching words and the conll output as an additional column (`sent`)
- `both_pretty`: same as above but instead of the raw conll string converts the sentence to a human readable format

The `matches` version of the output is useful if we want to further analyze
the results of our queries, not just count how many matches we got.
This will be more closely illustrated below.


The kind of queries presented above work okay with the kind of toy files used
here as examples. However, if we have a larger text, this really not an
effective way of mining it. However, if you don't mind the query running for a
while, you can, of course, give this approach a go. 
For that kind cases, `ApplyConllFilter` uses a progress bar with an estimation
of the remaining time (the proggress bar itself is part of the `dplyr`
package).

As a larger example, consider querying the entire (English translation of the)
gospel of John[^john] with a similar filter. This set of sentences is included
with the package as a vector called `gospel_of_john_sentences`.
Note that since we are now switching to a different
kind of parser output, we must define the expected columns
again using the `options` function.

```{r}


AnotherFilter <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat", "^V", use_regex=T) %>% 
        FilterConllRows("feat", "VV", is_negative=T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("feat",c("NN","NNP"))  %>% 
        return
}

options("conll_cols" = c("tokenid","token","lemma","feat","none", "head", "dep"))
start.time <- Sys.time()
matched_sentences <- ApplyConllFilter(gospel_of_john_sentences, AnotherFilter)
show(Sys.time() - start.time)
show(length(matched_sentences))

```

<!-- TODO: a function called GetColFromConll etc -->

As you can see, on my thinkpad x220 that took around 3 seconds -- still not
terribly bad. To get the perspective: this version of the gospel of John has 
`r length(gospel_of_john_sentences)  %>% format(big.mark=",")` sentences.
Of these, our search criteria were met by `r length(matched_sentences)`.

As a somewhat more serious work load, let's try to parse
the whole New testament. That means 9,329 sentences 
(not included in this package).

The result (on the same icore5 2.5ghz processor): `Time difference of 16.20708 secs`

So, let's say that 17 seconds for around 10,000 sentences. If it means
that 100,000 sentences will be analyzed in 170 seconds = a little less than
3 minutes, well -- you get the perspective. Not the most effective
way, but suitable for many cases, and if you've got time to wait,
even for some more serious text mining assignments.

[^john]: Included, also, as a sample file from the public domain World English
version and parsed with the Stanford coreNLP parser.


### Getting parameters from the results

In the previous examples we retrieved sentences by different
filters and so far we have relied on the raw output
of `ApplyConllFilter.` Let's move on to imagining a scenario in which
we want to get specific parameters out of the filtered results,
that is, to specifying `return_type = 'matches'`..

Consider the previous example with the sentences from the Gospel 
of John. Our little filter, repeated below, obtained all
the cases where a finite verb had a subject dependent which
was a noun:

```{r}


AnotherFilter <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat", "^V", use_regex=T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("feat",c("NN","NNP"))  %>% 
        return
}


matched_words <- ApplyConllFilter(gospel_of_john_sentences, AnotherFilter, "matches")

```

Now we have a tibble called `matched_words` which contains all the words (all 
the rows from the conll representations of sentences). With this tibble we
can do any statistical operations we would normally do with or data sets.
As an especially simple example, think about taking all the subject
dependents of finite verbs in the gospel of John and 
looking at which lemmas are most frequent as the subject:

```{r}

matched_words %>% count(lemma)  %>% arrange(desc(n))

```

No surprise, the most frequent subject is the proper name "Jesus".
But what is it that he keeps on doing? Let's modify
the filter a little:

```{r}



YetAnotherFilter <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat", "^V", use_regex=T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("lemma", "Jesus")  %>% 
        GetHeads(sentence) %>% 
        return
}

verbs_with_Jesus <- ApplyConllFilter(gospel_of_john_sentences,YetAnotherFilter,"both_pretty")

```

The filter became a little circular (we're first taking all the dependents
of verbs, then filtering those and then getting their heads again...),
but it does the job:

```{r}

verbs_with_Jesus %>% count(lemma) %>% arrange(desc(n))

```

Now, what if we would like to combine this information to get 
ngrams of sorts?

```{r}

a <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat", "^V", use_regex=T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("feat",c("NN","NNP"))  %>% 
        return
}


b <- function(sentence) {
    sentence  %>% 
        FilterConllRows("feat", "^V", use_regex=T) %>% 
        GetDeps(sentence) %>% 
        FilterConllRows("dep","nsubj") %>% 
        FilterConllRows("feat",c("NN","NNP"))  %>% 
        GetHeads(sentence) %>% 
        return
}


subj <- ApplyConllFilter(gospel_of_john_sentences,a,"both_pretty")
verb <- ApplyConllFilter(gospel_of_john_sentences,b,"both_pretty")

subj %>% 
    select(lemma, sent) %>% 
    left_join(verb  %>% select(lemma, sent), by="sent")   %>% 
    count(lemma.x, lemma.y) %>% 
    arrange(lemma.x, desc(n)) %>% 
    filter(n>4) %>% 
    print(n=99)

```

And we could, of course keep on going and draw
some graphs, build statistical models and so on...





## Filtering concordances

Although it is certainly possible to use this library for 
querying entire texts outputted in the conll format, the use case
for which the library was created is when you have a data frame
containing a list of concordances with more information than
just the actual conll-annotated sentence.

The basic idea is to have the data in a data frame / tibble where one row
corresponds to one sentence and there is a column containing the conll
representation of the sentence. In addition, you probably have some variables,
at least metadata an the like. One sample use case is, for instance, if you've queried 
for Twitter data and would like to have each tweet parsed and then use
the conll output of the tweet together with the massive amounts of 
metadata variables provided by the Twitter API for each tweet.

Here's a toy example of such a task.

Let's imagine we want to analyze the usage of the word "referee" in 
recent Uefa champions' league games (recent, because the Twitter api
only let's you search for tweets not older than one week).
In order to run the following you need to have requested 
developer access from Twitter, but never mind,
the data is also available as a small data set accompanying this 
package.

```{r}

#library(rtweet)
#ref <- search_tweets("(#UCL OR #ChampionsLeague OR #UEFAChampionsLeague) referee",  n = 5000)

```

I used the method described in the [webcorpcrawler package](https://github.com/utacorpora/webcorpcrawler/blob/master/docs/add_annotations_example.md)
to add syntactic annotations using Stanford Core-NLP. This gave me the following data set,
available as part of this package.
(I've stripped away most of the props offered by the Twitter API since there is A LOT of them)

```{r}

ucl_ref

```





such as the source of the concordance


Another basic use case is if you have a data set consisting of concordances,
acquired through, for instance, a query to a web corpus




# TODO

TODO: first make a reference data frame with all data as tabbed?
TODO: function to list possible columns and values






